Face Recognition: Too Bias, or Not Too Bias?

### 2020-07-09 cvprworkshop.tex
camera-ready

\usepackage{pifont}% http://ctan.org/pkg/pifont

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=.8\linewidth]{figures/montage.pdf}
    \caption{\small{\textbf{\gls{bfw}.} Average face of the different subsets: \emph{top-left}: the entire \gls{bfw}; \emph{top-row} per race;  \emph{left-column}: per gender. The others represent the ethnicity and gender of the race and gender, respectively. Table~\ref{tab:ethnic-splits} defines the acronyms of subgroups.}}
    \label{fig:avg-faces}
\end{figure}

@@

\begin{figure}[t!]
       \centering
    \includegraphics[width=.95\linewidth]{figures/fpr_percent_diff-crop.pdf}
    \caption{\small{\textbf{Percent difference from intended \gls{fpr}.} \emph{Top:} $t_g$ yields \gls{fpr} the span as large as 2$\times$ (\ie 200\%) that intended (\ie \gls{wm} for 1e-4). Furthermore, \gls{f} subgroups tend to perform worse than intended for all cases (while \gls{m} tend to overshoot intended performance, with exception of \gls{im} in for \gls{fpr}=1e-4). \emph{Bottom:} Subgroup-specific thresholds reduces this difference to near zero, where there are small differences, the percent difference across different subgroups is fair (\ie FPR=1e-4).}}\label{fig:percent:difference}
\end{figure}

@@

\begin{figure}[t!] 
	\centering    
	\includegraphics[width=.55\linewidth] {figures/human_eval.pdf}
		\caption{\small{\textbf{Human assessment (qualitative).} $\checkmark$ for \emph{match}; $\times$ for \emph{non-match}. Accuracy scores shown as bar plots. Humans are most successful at recognizing their own subgroup, with a few exceptions (\eg bottom).}}
		\label{fig:human-eval} 
\end{figure}

\glsresetall
\section{Conclusion}
We introduce a new data set \gls{bfw} with eight subgroups balanced across gender and ethnicity. With this, and upon highlighting the challenges and shortcomings of grouping subjects as a single subset, we provide evidence that forming subgroups is meaningful, as the FR algorithm rarely makes mistakes across subgroups. We used an \textit{off-the-shelf} Sphereface, hypothesizing this SOTA CNN suffers from bias because of the imbalanced train-set. Once established that the results do suffer from problems of bias, we observed that the same threshold across ethnic and gender subgroups leads to differences in the \gls{fpr} up to a factor of two. Also, the percent difference that seemingly the cause of the frenzy about bias in FR in main-stream media. Furthermore, we ameliorate these differences with a per-subgroup threshold, leveling out \gls{fpr}, and achieving a higher \gls{tpr}. We hypothesized that most humans grew among more than their own demographic and, therefore, effectively learn from imbalanced datasets-- a human evaluation validated that humans are biased, as most recognized their personal demographic best. The focused research findings presented here, along with the public database included, are extendable in vast ways. Thus, we see this as the slither to a much larger problem of bias in ML.

{\scriptsize
\bibliographystyle{ieee_fullname}
\bibliography{references}
}

>>>>

\usepackage{pifont}% http://ctan.org/pkg/pifont
\usepackage{balance}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=.8\linewidth]{figures/montage.pdf}
    \caption{\small{\textbf{\gls{bfw}.} Average face per subgroup: \emph{top-left}: the entire \gls{bfw}; \emph{top-row} per ethnicity;  \emph{left-column}: per gender. The others represent the ethnicity and gender, respectively. Table~\ref{tab:ethnic-splits} defines the acronyms of subgroups.}}
    \label{fig:avg-faces}
\end{figure}

@@

\begin{figure}[t!]
       \centering
    \includegraphics[width=.95\linewidth]{figures/fpr_percent_diff-crop.pdf}
    \caption{\small{\textbf{Percent difference from intended \gls{fpr}.} \emph{Top:} $t_g$ yields \gls{fpr} the span as large as 2$\times$ (\ie 200\%) that intended (\ie \gls{wm} for 1e-4). Furthermore, \gls{f} subgroups tend to perform worse than intended for all cases (while \gls{m} tend to overshoot intended performance, with exception of \gls{im} in for \gls{fpr}=1e-4). \emph{Bottom:} Subgroup-specific thresholds reduces this difference to near zero, where there are small differences, the percent difference across different subgroups is fair (\ie FPR=1e-4).}}\label{fig:percent:difference}
    \vspace{-4mm}
\end{figure}

@@

\begin{figure}[t!] 
	\centering    
	\includegraphics[width=.55\linewidth] {figures/human_eval.pdf}
		\caption{\small{\textbf{Human assessment (qualitative).} $\checkmark$ for \emph{match}; $\times$ for \emph{non-match}. Accuracy scores shown as bar plots. Humans are most successful at recognizing their own subgroup, with a few exceptions (\eg bottom).}}
		\label{fig:human-eval} 
		 \vspace{-5mm}
\end{figure}

\glsreset{bfw}
\section{Conclusion}
We introduce the \gls{bfw} dataset with eight subgroups balanced across gender and ethnicity. With this, and upon highlighting the challenges and shortcomings of grouping subjects as a single subset, we provide evidence that forming subgroups is meaningful, as the FR algorithm rarely makes mistakes across subgroups. We used an \textit{off-the-shelf} Sphereface, hypothesizing this SOTA CNN suffers from bias because of the imbalanced train-set. Once established that the results do suffer from problems of bias, we observed that the same threshold across ethnic and gender subgroups leads to differences in the \gls{fpr} up to a factor of two. Also, we clearly showed notable percent differences in ratings across subgroups. Furthermore, we ameliorate these differences with a per-subgroup threshold, leveling out \gls{fpr}, and achieving a higher \gls{tpr}. We hypothesized that most humans grew among more than their own demographic and, therefore, effectively learn from imbalanced datasets. In essence, a human evaluation validated that humans are biased, as most recognized their personal demographic best. This research, along with the data and resources, are extendable in vast ways. Thus, this is just a sliver of the larger problem of bias in ML.

{\small
\balance
\bibliographystyle{ieee_fullname}
\bibliography{references}
}

### 2020-07-09 sections/analysis.tex
camera-ready

\section{Results and Analysis}
A single \gls{cnn} was used as a means to control the experiments. For this, Sphereface~\cite{liu2017sphereface} trained on CASIA-Web~\cite{yi2014learning}, and evaluated on \gls{lfw}~\cite{LFWTech} (\%99.22 accuracy), encoded all of the faces.\footnote{\href{$https://github.com/clcarwin/sphereface\_pytorch$}{https://github.com/clcarwin/sphereface\_pytorch}} As reported in~\cite{wang2019racial}, \gls{lfw} has about a 13\%, 14\%, 3\%, and 70\% ratio in Asian, Black, Indian, and White, respectively. Furthermore, CASIA-Web is even more unbalanced (again, as reported in~\cite{wang2019racial}), with about  3\%, 11\%, 2\%, and 85\% for the same subgroups.
%\vspace{-4mm}
\glsunset{det}
% \subsection{\gls{det} analysis}
\noindent\paragraph{\gls{det} analysis.}
\gls{det} curves (5-fold, averaged) show per-subgroup trade-offs (Fig.~\ref{fig:detcurves}). Note that \gls{m} performs better than \gls{f}, precisely as one would expect from the tails of score-distributions for \emph{genuine} pairs (Fig.~\ref{fig:detection-model}). \Gls{af} and \gls{if} perform the worst.
\glsunset{m}
\glsunset{f}

@@

\glsunset{am}\glsunset{af}\glsunset{bm}\glsunset{bf}\glsunset{im}\glsunset{if}\glsunset{wm}\glsunset{wf}
\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{.3\linewidth}
    \includegraphics[width=.71\linewidth]{figures/curve_vgg16_subgroups-crop.pdf}
    \caption{VGG16~\cite{simonyan2014very}}
 \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_resnet50_subgroups-crop.pdf}
    \caption{ResNet50~\cite{he2016deep}}
   \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_senet50_subgroups-crop.pdf}
    \caption{SENet~\cite{hu2018squeeze}}
    \end{subfigure}
    \caption{\small{\textbf{\gls{det} curves for different CNNs}. \gls{fnr} (\%) (vertical) vs \gls{fpr}  (horizontal, log-scale) for VGG2~\cite{Cao18} models with different backbones (VGG16, Resnet50, SEnet50). Lower is better. For each plot, \gls{wm} is the top-performer, while \gls{af} is the worst. The ordering of the curves is roughly the same for each backbone.}}\label{fig:sdm-appendix-a}
%    \vspace{-5mm}
\end{figure*}
\noindent\paragraph{Score analysis.}
% \subsection{Score analysis}
Fig.~\ref{fig:detection-model} shows score distributions for faces of the same (\ie \emph{Genuine}) and different (\ie \emph{Imposter}) identity, with a subgroup per \gls{sdm} plot. Notice that score distributions for imposters tend to peak about zero for all subgroups, and with minimal deviation comparing modes of the different plots. On the other hand, the score distribution of the \emph{genuine} pairs varies across subgroups in location (\ie score value) and spread (\ie overall shape). Asian Fem
Fig.~\ref{fig:confusion} shows the confusion matrix of the subgroups. A vast majority of errors occur in the intra-subgroup. It is interesting to note that while the definition of  each group  based on ethnicity and race may not be crisply defined, the confusion matrix indicates that in practice, the \gls{cnn} finds that the groups are effectively separate. The categories are, therefore, meaningful for \gls{fr}.

@@

\begin{table}[b!]
%    \vspace{-4mm}
\glsunset{tar}
\glsunset{far}
\caption{\small{\textbf{\gls{tar} at intervals of \gls{far}}. \gls{far}, listed are the \gls{tar} scores for a global threshold (top) and the proposed category-based threshold (bottom). Higher is better.}}\label{tab:ethnicy-far} 
% \vspace{-2mm}
\begin{center}
\scriptsize
\begin{tabular}{l c c c c c}
     \gls{far} & 0.3 & 0.1 & 0.01 & 0.001 & 0.0001\\\midrule
    \multirow{2}{.1mm}{\textbf{\gls{af}}} &0.990 & 0.867 & 0.516 & 0.470 & 0.465\\[-4pt]
        &1.000 & 0.882 & 0.524 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{am}}} &0.994 & 0.883 & 0.529 & 0.482 & 0.477\\[-4pt]
        &1.000 & 0.890 & 0.533 & 0.486 & 0.482\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bf}}} &0.991 & 0.870 & 0.524 & 0.479 & 0.473\\[-4pt]
        &1.000 & 0.879 & 0.530 & 0.484 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bm}}} &0.992 & 0.881 & 0.526 & 0.480 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.532 & 0.485 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{if}}} &0.996 & 0.881 & 0.532 & 0.486 & 0.481\\[-4pt]
        &1.000 & 0.884 & 0.534 & 0.488 & 0.484\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{im}}} &0.997 & 0.895 & 0.533 & 0.485 & 0.479\\[-4pt]
        &1.000 & 0.898 & 0.535 & 0.486 & 0.481\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wf}}} &0.988 & 0.878 & 0.517 & 0.469 & 0.464\\[-4pt]
        &1.000 & 0.894 & 0.526 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wm}}} &0.989 & 0.896 & 0.527 & 0.476 & 0.470\\[-4pt]
        &1.000 & 0.910 & 0.535 & 0.483 & 0.478\\[-1pt]
    \midrule
    \multirow{2}{3mm}{\textbf{Avg.}} &0.992 & 0.881 & 0.526 & 0.478 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.531 & 0.483 & 0.479\\[-10pt]
\end{tabular}
\end{center}
\glsreset{tar}
\glsreset{far}
%  \vspace{-3mm}

@@

To further demonstrate the extent of the problem, we follow settings typical for systems in practice. We set the desired \gls{fpr}, and then determine the percent difference, \ie desired versus actual (Fig.~\ref{fig:percent:difference}, \emph{top}). Furthermore, we mitigate this highly skewed phenomena by applying subgroup-specific thresholds (\emph{bottom}). By this, minimal error from the desired \gls{fpr}. In addition, where there is small error, the offset is balanced across subgroups.

% \subsubsection{Other \gls{cnn} models.}~\label{app:sec:other:models}
%\vspace{-5mm}
\noindent\paragraph{Model analysis.}
Variations in optimal threshold exist across models (Fig.~\ref{fig:sdm-appendix-a}). Like in Fig.~\ref{fig:detcurves}, the \gls{det} curves for three \gls{cnn}-based models, each trained on VGG2 with softmax but with different backbones.\footnote{\href{https://github.com/rcmalli/keras-vggface}{https://github.com/rcmalli/keras-vggface}} Notice similar trends across subgroups and models, which is consistent with  Sphereface as well (Fig.~\ref{fig:detcurves}). For example, the plots generated with Sphereface and VggFace2 all have the \gls{wm} curve at the bottom (\ie best) and \gls{af} on top (\ie worst). Thus, the additional \gls{cnn}-based models demonstrate the same phenomena: proportional to the overall model performance, exact in the ordering of curves for each subgroup.

%\vspace{-5mm}
\noindent\paragraph{Verification threshold analysis.}%\subsection{Verification threshold} \label{subsec:analysis:verification}
We seek to reduce the bias between subgroups. Such that an operating point (\ie \gls{fpr}) is constant across subgroups. To accomplish that, we used a per subgroup threshold. In \gls{fv}, we consider one image as the query, and all others as the test. For this, the ethnicity of the query image is assumed. We can then examine the \gls{det} curves and pick the best threshold per subgroup for a certain \gls{fpr}.

@@

        &WF & 44.7 &43.8 & \textbf{\textcolor{blue}{57.3}}& \textbf{\textcolor{red}{48.0}} & \textbf{48.5} \\
        &WM & 30.1&\textbf{\textcolor{red}{47.4}} &  45.3 & \textbf{\textcolor{blue}{56.1}} & 44.7\\\cline{3-7}
        &Avg &  43.3& 47.4&\textbf{47.7} &46.3 &46.2\\
 \end{tabular}
 \end{tabular}
 }
 \end{center}
% \vspace{-7mm}
\end{table}

%\vspace{-5mm}
\noindent\paragraph{Human evaluation analysis.}
% \subsection{Human evaluation}
Subjects of a subgroup likely have mostly been exposed to others of the same (Table~\ref{tab:humsn-eval-results} and Fig.~\ref{fig:human-eval}). Therefore, it is expected they would be best at labeling their own; similar for the same ethnicity, but another gender. Our findings concur. Each subgroup is best at labeling their type, and then second best at labeling the same ethnicity but opposite sex. Interestingly, each group of images is best tagged by the corresponding subgroup, with the second-best having the same ethnicity and opposite gender. On average, subgroups are comparable at labeling images. Analogous to the \gls{fr} system, performance ratings differ when analyzing within and across subgroups. In other words, performance on \gls{bfw} improved with subgroup-specific thresholds. Similarly, humans tend to better recognize individuals by facial cues of the same or similar demographics. Put differently, as the recognition performances drop with a global threshold optimized for one subgroup and deployed on another, human performance tends to drop when across subgroups (\ie performances drop for less familiar subgroups).

>>>>

\section{Results and Analysis}
A single \gls{cnn} was used as a means to control the experiments. For this, Sphereface~\cite{liu2017sphereface} trained on CASIA-Web~\cite{yi2014learning}, and evaluated on \gls{lfw}~\cite{LFWTech} (\%99.22 accuracy), encoded all of the faces.\footnote{\href{$https://github.com/clcarwin/sphereface\_pytorch$}{https://github.com/clcarwin/sphereface\_pytorch}} As reported in~\cite{wang2019racial}, \gls{lfw} has about 13\%, 14\%, 3\%, and 70\% ratio in Asian, Black, Indian, and White, respectively. Furthermore, CASIA-Web is even more unbalanced (again, as reported in~\cite{wang2018racial}), with about  3\%, 11\%, 2\%, and 85\% for the same subgroups.
%\vspace{-4mm}
\glsunset{det}
% \subsection{\gls{det} analysis}
\vspace{-5mm}
\noindent\paragraph{\gls{det} analysis.}
\gls{det} curves (5-fold, averaged) show per-subgroup trade-offs (Fig.~\ref{fig:detcurves}). Note that \gls{m} performs better than \gls{f}, precisely as one would expect from the tails of score-distributions for \emph{genuine} pairs (Fig.~\ref{fig:detection-model}). \Gls{af} and \gls{if} perform the worst.
\glsunset{m}
\glsunset{f}

@@

\glsunset{am}\glsunset{af}\glsunset{bm}\glsunset{bf}\glsunset{im}\glsunset{if}\glsunset{wm}\glsunset{wf}
\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{.3\linewidth}
    \includegraphics[width=.71\linewidth]{figures/curve_vgg16_subgroups-crop.pdf}
    \caption{VGG16~\cite{simonyan2014very}}
 \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_resnet50_subgroups-crop.pdf}
    \caption{ResNet50~\cite{he2016deep}}
   \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_senet50_subgroups-crop.pdf}
    \caption{SENet~\cite{hu2018squeeze}}
    \end{subfigure}
    \caption{\small{\textbf{\gls{det} curves for different CNNs}. \gls{fnr} (\%) (vertical) vs \gls{fpr}  (horizontal, log-scale) for VGG2~\cite{Cao18} models with different backbones (VGG16, Resnet50, SEnet50). Lower is better. For each plot, \gls{wm} is the top-performer, while \gls{af} is the worst. The ordering of the curves is roughly the same for each backbone.}}\label{fig:sdm-appendix-a}
    \vspace{-5mm}
\end{figure*}

\vspace{-5mm}
\noindent\paragraph{Score analysis.}
% \subsection{Score analysis}
Fig.~\ref{fig:detection-model} shows score distributions for faces of the same (\ie \emph{genuine}) and different (\ie \emph{imposter}) identity, with a subgroup per \gls{sdm} graph. Notice that score distributions for imposters tend to peak about zero for all subgroups, and with minimal deviation comparing modes of the different plots. On the other hand, the score distribution of the \emph{genuine} pairs varies across subgroups in location (\ie score value) and spread (\ie overall shape).
Fig.~\ref{fig:confusion} shows the confusion matrix of the subgroups. A vast majority of errors occur in the intra-subgroup. It is interesting to note that while the definition of  each group  based on ethnicity and race may not be crisply defined, the confusion matrix indicates that in practice, the \gls{cnn} finds that the groups are effectively separate. The categories are, therefore, meaningful for \gls{fr}.

@@

\begin{table}[b!]
    \vspace{-4mm}
\glsunset{tar}
\glsunset{far}
\caption{\small{\textbf{\gls{tar} at intervals of \gls{far}}. \gls{far}, listed are the \gls{tar} scores for a global threshold (top) and the proposed category-based threshold (bottom). Higher is better.}}\label{tab:ethnicy-far} 
\vspace{-2mm}
\begin{center}
\scriptsize
\begin{tabular}{l c c c c c}
     \gls{far} & 0.3 & 0.1 & 0.01 & 0.001 & 0.0001\\\midrule
    \multirow{2}{.1mm}{\textbf{\gls{af}}} &0.990 & 0.867 & 0.516 & 0.470 & 0.465\\[-4pt]
        &1.000 & 0.882 & 0.524 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{am}}} &0.994 & 0.883 & 0.529 & 0.482 & 0.477\\[-4pt]
        &1.000 & 0.890 & 0.533 & 0.486 & 0.482\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bf}}} &0.991 & 0.870 & 0.524 & 0.479 & 0.473\\[-4pt]
        &1.000 & 0.879 & 0.530 & 0.484 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bm}}} &0.992 & 0.881 & 0.526 & 0.480 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.532 & 0.485 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{if}}} &0.996 & 0.881 & 0.532 & 0.486 & 0.481\\[-4pt]
        &1.000 & 0.884 & 0.534 & 0.488 & 0.484\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{im}}} &0.997 & 0.895 & 0.533 & 0.485 & 0.479\\[-4pt]
        &1.000 & 0.898 & 0.535 & 0.486 & 0.481\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wf}}} &0.988 & 0.878 & 0.517 & 0.469 & 0.464\\[-4pt]
        &1.000 & 0.894 & 0.526 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wm}}} &0.989 & 0.896 & 0.527 & 0.476 & 0.470\\[-4pt]
        &1.000 & 0.910 & 0.535 & 0.483 & 0.478\\[-1pt]
    \midrule
    \multirow{2}{3mm}{\textbf{Avg.}} &0.992 & 0.881 & 0.526 & 0.478 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.531 & 0.483 & 0.479\\[-10pt]
\end{tabular}
\end{center}
\glsreset{tar}
\glsreset{far}
 \vspace{-3mm}

@@

To further demonstrate the extent of the problem, we follow settings typical for systems in practice. We set the desired \gls{fpr}, and then determine the percent difference, \ie desired versus actual (Fig.~\ref{fig:percent:difference}, \emph{top}). Furthermore, we mitigate this highly skewed phenomenon by applying subgroup-specific thresholds (\emph{bottom}) - by this, minimal error from the desired \gls{fpr}. Besides where there is a small error, the offset is balanced across subgroups.

% \subsubsection{Other \gls{cnn} models.}~\label{app:sec:other:models}
\vspace{-5mm}
\noindent\paragraph{Model analysis.}
Variations in optimal threshold exist across models (Fig.~\ref{fig:sdm-appendix-a}). Like in Fig.~\ref{fig:detcurves}, the \gls{det} curves for three \gls{cnn}-based models, each trained on VGG2 with softmax but with different backbones.\footnote{\href{https://github.com/rcmalli/keras-vggface}{https://github.com/rcmalli/keras-vggface}} Notice similar trends across subgroups and models, which is consistent with  Sphereface as well (Fig.~\ref{fig:detcurves}). For example, the plots generated with Sphereface and VggFace2 all have the \gls{wm} curve at the bottom (\ie best) and \gls{af} on top (\ie worst). Thus, the additional \gls{cnn}-based models demonstrate the same phenomena: proportional to the overall model performance, exact in the ordering of curves for each subgroup.

\vspace{-5mm}
\noindent\paragraph{Verification threshold analysis.}%\subsection{Verification threshold} \label{subsec:analysis:verification}
We seek to reduce the bias between subgroups. Such that an operating point (\ie \gls{fpr}) is constant across subgroups. To accomplish that, we used a per subgroup threshold. In \gls{fv}, we consider one image as the query, and all others as the test. For this, the ethnicity of the query image is assumed. We can then examine the \gls{det} curves and pick the best threshold per subgroup for a certain \gls{fpr}.

@@

        &WF & 44.7 &43.8 & \textbf{\textcolor{blue}{57.3}}& \textbf{\textcolor{red}{48.0}} & \textbf{48.5} \\
        &WM & 30.1&\textbf{\textcolor{red}{47.4}} &  45.3 & \textbf{\textcolor{blue}{56.1}} & 44.7\\\cline{3-7}
        &Avg &  43.3& 47.4&\textbf{47.7} &46.3 &46.2\\
 \end{tabular}
 \end{tabular}
 }
 \end{center}
\vspace{-5mm}
\end{table}

\vspace{-5mm}
\noindent\paragraph{Human evaluation analysis.}
% \subsection{Human evaluation}
Subjects of a subgroup likely have mostly been exposed to others of the same (Table~\ref{tab:humsn-eval-results} and Fig.~\ref{fig:human-eval}). Therefore, it is expected they would be best at labeling their own, similar to the same ethnicity, but another gender. Our findings concur. Each subgroup is best at labeling their type, and then second best at labeling the same ethnicity but opposite sex.

Interestingly, each group of images is best tagged by the corresponding subgroup, with the second-best having the same ethnicity and opposite gender. On average, subgroups are comparable at labeling images. Analogous to the \gls{fr} system, performance ratings differ when analyzing within and across subgroups. In other words, performance on \gls{bfw} improved with subgroup-specific thresholds. Similarly, humans tend to better recognize individuals by facial cues of the same or similar demographics. Put differently, as the recognition performances drop with a global threshold optimized for one subgroup and deployed on another, human performance tends to fall when across subgroups (\ie performances drop for less familiar subgroups).

### 2020-07-09 sections/introduction.tex
camera-ready

\glsresetall
\section{Introduction}
    As more of society becomes integrated with \gls{ml}, topics related to bias, fairness, and the formalization of \gls{ml} standards attract more attention~\cite{10.1007/978-3-030-13469-3_68, anne2018women, wang2018racial}. Thus, an effect of the growing dependency on \gls{ml} is a growing concern for biased and unfair algorithms, \eg untrustworthy and racist \gls{fr} systems~\cite{england2019, nagpal2019deep,snow2018}.

    Nowadays, \glspl{cnn} are trained on faces identified by a detection system. Specifically, for \gls{fr}, the goal is to encode faces in an N-dimensional space such that samples of the same identity are neighbors, while different identities are farther apart. Thus, we can deploy a \gls{cnn} to encode faces, and then compare via a similarity score (\ie pairs with a high enough score classified as \emph{genuine}; else, \emph{imposter}).

@@

Typically, a fixed threshold sets the decision boundary by which to compare scores (Fig.~\ref{fig:bio-metrics}). Thus, features of the same identity must satisfy a criterion based on a single value~\cite{deng2019arcface, liu2017sphereface, wang2018additive, wang2018cosface}. However, we found that a single (\ie global) threshold is a crude measure that leads to skewed errors. Furthermore, the held-out set used to determine the threshold tends to share the same distribution with the test set, favoring specific demographics that are the majority. That skew (\ie the difference in the performance of an algorithm of certain demographics) is our definition of bias. A key question is: \emph{is \gls{fr} too biased, or not?}

\begin{table*}[!t]
    \centering
    \caption{\small{\textbf{Database stats and nomenclature, optimal thresholds ($t_o$), and accuracy scores.} \textit{Header:} Subgroup definitions. \textit{Top:} Statistics of \gls{bfw}. \textit{Bottom::} Number of pairs for each partition. Columns grouped by race and then further split by gender. Out of millions of pairs, accuracy is inconsistent across subgroups. Furthermore, $F$ tends to perform inferior to that of $M$ for all but B.}}\label{tab:ethnic-splits}
    \scalebox{.75}{
     \resizebox{\textwidth}{!}{%
    \begin{tabular}{r c c c c c c c c l}
        \toprule
        & \multicolumn{2}{c}{Asian (A)} & \multicolumn{2}{c}{Black (B)}  & \multicolumn{2}{c}{Indian (I)}& \multicolumn{2}{c}{White (W)}\\
        \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7}\cmidrule(l){8-9} % spanning less than the full width of the table - you can add (r) or (l) just before the opening curly bracket to shorten the rule on the left or right side
         & Female (AF) & Male (AM) & BF & BM& IF & IM & WF & WM&Aggregated\\ % Column names row
        \midrule

@@

        % Acc$@t_o$ & 0.942 &0.952 &0.968 &0.961 &0.960 &0.962 & 0.974 &0.982 & 0.963 $\pm$ 0.012\\
        \bottomrule
    \end{tabular}}
    }
    % \vspace{-12pt}
\end{table*}

    Making matters more challenging is that race and ethnicity are loosely defined. For example, the US Census Bureau allows an individual to self-identify race.\footnote{\scriptsize\href{https://www.census.gov/mso/www/training/pdf/race-ethnicity-onepager.pdf}{www.census.gov/mso/www/training/pdf/race-ethnicity-onepager}} For this work, we define subgroups as specific sub-populations with face characteristics similar to those found in a region.

    The adverse effects of a global threshold are two-fold: \textbf{(1)} the mapping produced by a \gls{cnn} is nonuniform. Therefore, distances between pairs of faces in different demographics vary in distribution of similarity scores (Fig~\ref{fig:detection-model}); \textbf{(2)} the evaluation set is imbalanced as well. Particular demographics making up a majority of the population will carry most weight on the reported performance ratings. Reported results favor the common traits over the underrepresented. Alas, demographics like gender, ethnicity, race, and age are underrepresented in most public datasets~\cite{merler2019diversity, wang2018racial}.

    For \textbf{(1)} we propose subgroup-specific (\ie optimal) thresholds, while we address \textbf{(2)} with a new benchmark dataset for measuring bias in \gls{fr}, \gls{bfw} (Table~\ref{tab:ethnic-splits} and~\ref{tab:compared}). \gls{bfw} allows for a fair evaluation of \gls{fr} systems, while enabling demographic-specific ratings to be reported. We use \gls{bfw} to gain a deeper understanding of the extent to which bias presents in \gls{soa} \gls{cnn} used to encode faces. Then, we suggest a mechanism to mitigate problems of bias with more balanced performance ratings for different demographics. Specifically, we propose using an adaptive threshold that varies depending on the characteristics of detected facial attributes (\ie gender and ethnicity, Fig~\ref{fig:avg-faces}). We show an increase in accuracy with a balanced performance for different subgroups. Similarly, we show a positive effect of adjusting the similarity threshold based on the facial features of matched faces. Thus, selective use of similarity thresholds in current \gls{soa} \gls{fr} systems provides more intuition in \gls{fr} research with a method easy to adopt in practice.

>>>>

\glsresetall
\section{Introduction}
    As more of society becomes integrated with \gls{ml}, topics related to bias, fairness, and the formalization of \gls{ml} standards attract more attention~\cite{10.1007/978-3-030-13469-3_68, anne2018women, wang2018racial}. Thus, an effect of the growing dependency on \gls{ml} is an ever-increasing concern about the biased and unfair algorithms, \eg untrustworthy and prejudiced \gls{fr}~\cite{nagpal2019deep,snow2018}.

    Nowadays, \glspl{cnn} are trained on faces identified by a detection system. Specifically, for \gls{fr}, the goal is to encode faces in an N-dimensional space such that samples of the same identity are neighbors, while different persons are farther apart. Thus, we can deploy a \gls{cnn} to encode faces, and then compare via a similarity score (\ie pairs with a high enough score classified as \emph{genuine}; else, \emph{imposter}).

@@

Typically, a fixed threshold sets the decision boundary by which to compare scores (Fig.~\ref{fig:biometrics}). As such, features of the same identity must satisfy a criterion based on a single value~\cite{deng2019arcface, liu2017sphereface, wang2018additive, wang2018cosface}. However, we found that an individual (\ie global) threshold is a crude measure that leads to skewed errors. Furthermore, the held-out set used to determine the threshold tends to share the same distribution with the test set, favoring specific demographics that are the majority. That skew (\ie the difference in the performance of an algorithm of particular demographics) is our definition of bias. A key question is: \emph{is \gls{fr} too biased, or not?}

\begin{table*}[!t]
    \centering
    \caption{\small{\textbf{Database stats and nomenclature.} \textit{Header:} Subgroup definitions. \textit{Top:} Statistics of \gls{bfw}. \textit{Bottom:} Number of pairs for each partition. Columns grouped by ethnicity and then further split by gender.}}\label{tab:ethnic-splits}
    \scalebox{.75}{
     \resizebox{\textwidth}{!}{%
    \begin{tabular}{r c c c c c c c c l}
        \toprule
        & \multicolumn{2}{c}{Asian (A)} & \multicolumn{2}{c}{Black (B)}  & \multicolumn{2}{c}{Indian (I)}& \multicolumn{2}{c}{White (W)}\\
        \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7}\cmidrule(l){8-9} % spanning less than the full width of the table - you can add (r) or (l) just before the opening curly bracket to shorten the rule on the left or right side
         & Female (AF) & Male (AM) & BF & BM& IF & IM & WF & WM&Aggregated\\ % Column names row
        \midrule

@@

        % Acc$@t_o$ & 0.942 &0.952 &0.968 &0.961 &0.960 &0.962 & 0.974 &0.982 & 0.963 $\pm$ 0.012\\
        \bottomrule
    \end{tabular}}
    }
    \vspace{-12pt}
\end{table*}

    Making matters more challenging, precise definition of race and ethnicity vary from source-to-source. For example, the US Census Bureau allows an individual to self-identify race.\footnote{\scriptsize\href{https://www.census.gov/mso/www/training/pdf/race-ethnicity-onepager.pdf}{www.census.gov/mso/www/training/pdf/race-ethnicity-onepager}} Even gender, our attempt to encapsulate the complexities of the sex of a human as one of two labels. Others have addressed the oversimplified class labels by representing gender as a continuous value between 0 and 1 - rarely is a person entirely \emph{M} or \emph{F}, but most are somewhere in between~\cite{merler2019diversity}. For this work, we define subgroups as specific sub-populations with face characteristics similar to others in a region. Specifically, we focus on 8 subgroups (Fig.~\ref{fig:avg-faces}).

    The adverse effects of a global threshold are two-fold: \textbf{(1)} mappings produced by \glspl{cnn} are nonuniform. Therefore, distances between pairs of faces in different demographics vary in distribution of similarity scores (Fig~\ref{fig:detection-model}); \textbf{(2)} evaluation set is imbalanced. Subgroups that make up a majority of the population will carry most weight on the reported performance ratings. Reported results favor the common traits over the underrepresented. Demographics like gender, ethnicity, race, and age are underrepresented in most public datasets~\cite{merler2019diversity, wang2018racial}.

    For \textbf{(1)}, we propose subgroup-specific (\ie optimal) thresholds while addressing \textbf{(2)} with a new benchmark dataset to measure bias in \gls{fr}, \gls{bfw} (Table~\ref{tab:ethnic-splits} and~\ref{tab:compared}). \gls{bfw} serves as a proxy for fair evaluations for \gls{fr} while enabling per subgroup ratings to be reported. We use \gls{bfw} to gain an understanding of the extent to which bias is present in \gls{soa} \gls{cnn}s used \gls{fr}. Then, we suggest a mechanism to mitigate problems of bias with more balanced performance ratings for different demographics. Specifically, we propose using an adaptive threshold that varies depending on the characteristics of detected facial attributes (\ie gender and ethnicity). We show an increase in accuracy with a balanced performance for different subgroups. Similarly, we show a positive effect of adjusting the similarity threshold based on the facial features of matched faces. Thus, selective use of similarity thresholds in current \gls{soa} \gls{fr} systems provides more intuition in \gls{fr} research with a method easy to adopt in practice.
